\vspace{-\baselineskip} Fast and scalable Gaussian process modeling with applications to astronomical time series \vspace{-2\baselineskip}

### 2017-06-09 paper/ms.tex
some edits
typo

The growing field of large-scale time domain astronomy requires methods for
probabilistic data analysis that are computationally tractable, even with
large datasets.
Gaussian Processes are a popular class of models used for this purpose but,
since the computational cost scales as the cube of the number of data points,
their application has been limited to relatively small datasets.
In this paper, we present a method for Gaussian Process modeling in
one-dimension where the computational requirements scale linearly with the
size of the dataset.
We demonstrate the method by applying it to simulated and real astronomical
time series datasets.
These demonstrations are examples of probabilistic inference of stellar
rotation periods, asteroseismic oscillation spectra, and transiting planet
parameters.
The method exploits structure in the problem when the covariance function is
expressed as a mixture of complex exponentials, without requiring evenly
spaced observations or uniform noise.
This form of covariance arises naturally when the process is a mixture of
stochastically-driven damped harmonic oscillators~--~providing a physical
motivation for and interpretation of this choice~--~but we also demonstrate
that it is effective in other cases.
We present a mathematical description of the method, the details of the
implementation, and a comparison to existing scalable Gaussian Process methods.
The method is flexible, fast, and most importantly, interpretable, with a wide
range of potential applications within astronomical data analysis and beyond.
We provide well-tested and documented open-source implementations of this
method in \project{C++}, \project{Python}, and \project{Julia}.
A change.

@@

We now turn to a discussion of \celeriteterm\ as a model of astrophysical
variability.
A common concern in the context of GP modeling in astrophysics is the lack of
physical motivation for the choice of kernel functions.
Kernels are often chosen simply because they are popular, with little
consideration of the impact of this decision.
While this doesn't necessarily impact the inferred results, in this section we
discuss an exact physical interpretation of the \celeriteterm\ kernel that
will be applicable to many astrophysical systems but especially the
time-variability of stars.

@@

For these definitions, the kernel is
\begin{equation}\eqlabel{sho-kernel}
k_\mathrm{SHO}(\tau;\,S_0,\,Q,\,\omega_0) =
S_0\,\omega_0\,Q\,e^{-\frac{\omega_0\,\tau}{2Q}}\,
\begin{cases}
    \cosh{(\eta\,\omega_0\,\tau)} +
        \frac{1}{2\,\eta\,Q}\,\sinh{(\eta\,\omega_0\,\tau)}, & 0 < Q < 1/2\\
    2\,(1+\omega_0\,\tau), & Q = 1/2\\
    \cos{(\eta\,\omega_0\,\tau)} +
        \frac{1}{2\,\eta\,Q} \sin{(\eta\,\omega_0\,\tau)},& 1/2 < Q\\
\end{cases}
\end{equation}
where $\eta = \vert 1-(4\,Q^2)^{-1}\vert^{1/2}$.
We note that, because of the damping, the characteristic oscillation frequency
in this model, $d_j$, for any finite quality factor $Q > 1/2$, is not equal to
the frequency of the undamped oscillator, $\omega_0$.

@@

{\item For $Q = 1/\sqrt{2}$, \eq{sho-psd} simplifies to
\begin{eqnarray}\eqlabel{granulation-psd}
S(\omega) = \sqrt{\frac{2}{\pi}}\,\frac{S_0}{(\omega/\omega_0)^4+1} \quad.
\end{eqnarray}
This functional form is commonly used to model for the background granulation
noise in asteoreseismic and helioseismic \citep{Harvey:1985, Michel:2009,
Kallinger:2014}  analyses.
The kernel for this value of $Q$ is
\begin{equation}
k(\tau) = S_0\,\omega_0\,e^{-\frac{1}{\sqrt{2}}\,\omega_0\,\tau}\,
    \cos{\left(\frac{\omega_0\,\tau}{\sqrt{2}}-\frac{\pi}{4}\right)} \quad.
\end{equation}}

{\item Substituting $Q = 1/2$, \eq{sho-psd} becomes
\begin{eqnarray}
S(\omega) =
    \sqrt{\frac{2}{\pi}}\,\frac{S_0}{\left[(\omega/\omega_0)^2+1\right]^2}
\end{eqnarray}
with the corresponding covariance function (using
\eqalt{celerite-kernel} and \eqalt{sho-real})
\begin{eqnarray}\eqlabel{approx-matern}
k(\tau) &=& \lim_{f \to 0}\,
    \frac{1}{2}\,S_0\,\omega_0\,
    \left[\left(1+1/f\right)\,e^{-\omega_0\,(1-f)\,\tau} +
          \left(1-1/f\right)\,e^{-\omega_0\,(1+f)\,\tau}
    \right] \\
&=& S_0\,\omega_0\,e^{-\omega_0\,\tau}\,[1+\omega_0\,\tau]
\end{eqnarray}
or equivalently (using \eqalt{celerite-kernel} and \eqalt{sho-complex})
\begin{eqnarray}\eqlabel{approx-matern2}
k(\tau) &=& \lim_{f \to 0}\,
    S_0\,\omega_0\,e^{-\omega_0\,\tau}\,
    \left[\cos(f\,\tau) + \frac{\omega_0}{f}\,\sin(f\,\tau)\right] \\
&=& S_0\,\omega_0\,e^{-\omega_0\,\tau}\,[1+\omega_0\,\tau] \quad.
\end{eqnarray}
This covariance function is also known as the Mat\'ern-3/2 function
\citep{Rasmussen:2006}.
This suggests that the Mat\'ern-3/2 covariance can be approximated using
the \celeriteterm\ framework with a small value of $f$ in \eq{approx-matern2} but we
caution that this might lead to numerical issues for the solver.
}

@@

The relationship between the \celeriteterm\ model and semiseparable linear
algebra was first recognized by Ambikasaran \citep{Ambikasaran:2015} building
on earlier work by Rybicki \& Press \citep{Rybicki:1995}.
\citet{Ambikasaran:2015} derived a scalable direct solver for any general
semiseparable matrix and applied this method to \celeriteterm\ models with
$b_j = 0$ and $d_j = 0$.
During the preparation of this paper, we generalized this earlier method to
include non-zero $b_j$ and $d_j$, but it turns out that we can derive a higher
performance algorithm by restricting our method to positive semi-definite
(PSD) matrices.
In \app{psd}, we discuss methods for ensuring that a \celeriteterm\ model is
PSD.

@@

\subsection{Cholesky factorization of PSD semiseparable matrices}

For our purposes, we want to compute the log-determinant of $K$ and solve
linear systems of the form $K\,\bvec{z}=\bvec{y}$ for $\bvec{z}$, where $K$ is
a rank-$R$ semiseparable PSD matrix.
These can both be computed using the Cholesky factorization of $K$:
\begin{eqnarray}
K = L\,D\,L^\T
\end{eqnarray}
where $L$ is a lower triangular matrix with unit diagonal and $D$ is a
diagonal matrix.
We use the ansatz that $L$ has the form
\begin{eqnarray}
    L = I + \mathrm{tril} (U\,W^T)
\end{eqnarray}
where $I$ is the identity, $U$ is the matrix from above, and $W$ is an unknown
$N \times R$ matrix.
Combining this with \eq{semi-sep}, we find the equation
\begin{eqnarray}\eqlabel{cholesky}
A + \mathrm{tril}(U\,V^\T) + \mathrm{triu}(V\,U^\T)
&=& \left[I + \mathrm{tril} (U\,W^T)\right]\,D\,
    {\left[I + \mathrm{tril} (U\,W^T)\right]}^\T \\
&=& D + \mathrm{tril}(U\,W^\T)\,D
    + D\,\mathrm{triu}(W\,U^\T) \nonumber\\
&& +\, \mathrm{tril}(U\,W^\T)\,D\,\mathrm{triu}(W\,U^\T) \nonumber
\end{eqnarray}
that we can solve for $W$.
Setting each element on the left side of \eq{cholesky} equal to the
corresponding element on the right side, we derive the following recursion
relations
\begin{eqnarray}
S_{n,j,k} &=& S_{n-1,j,k} + D_{n-1,n-1}\,W_{n-1,j}\,W_{n-1,k} \nonumber\\
D_{n,n} &=& A_{n,n} -
    \sum_{j=1}^{R}\sum_{k=1}^{R} U_{n,j}\,S_{n,j,k}\,U_{n,k}
    \nonumber\\
W_{n,j} &=& \frac{1}{D_{n,n}}\left[ V_{n,j} -
    \sum_{k=1}^{R}U_{n,k}\,S_{n,j,k} \right]
    \eqlabel{algorithm}
\end{eqnarray}
where $S_n$ is a symmetric $R \times R$ matrix and every element
$S_{1,j,k}=0$.
The computational cost of one iteration of the update in \eq{algorithm} is
$\mathcal{O}(R^2)$ and this must be run $N$ times~--~once for each row~--~so
the total cost of this factorization scales as $\mathcal{O}(N\,R^2)$.

After computing the Cholesky factorization of this matrix $K$, we can also
apply its inverse and compute its log-determinant in $\mathcal{O}(N\,R)$ and
$\mathcal{O}(N)$ operations respectively.
The log-determinant of $K$ is given by
\begin{eqnarray}
    \ln \det K = \sum_{n=1}^N \ln D_{n,n}
\end{eqnarray}
where, since $K$ is PSD, $D_{n,n}>0$ for all $n$.

@@

The algorithm derived here is a general method for factorizing, solving, and
computing the determinant of PSD rank-$R$ semiseparable matrices.
As we will show below, with some care, this method can be used to compute
\eq{gp-likelihood} for \celeriteterm\ models.
In practice, when $K$ is PSD and this method is applicable, we find that this
method is about a factor of 20 faster than an optimized implementation of the
earlier, more general method from \citet{Ambikasaran:2015}.

@@

First, we empirically confirm that the method solves systems of the form
$K\,\bvec{z} = \bvec{y}$ to high numerical accuracy.
The left panel of \Figure{error} shows the L-infinity norm for the residual
$\bvec{z} - K^{-1}\,\bvec{y}$ for a range of dataset sizes $N$ and numbers of
terms $J$.
For each pair of $N$ and $J$ we repeated the experiment 10 times with
different randomly generated datasets and \celeriteterm\ parameters and
averaged across experiments to remove some sampling noise from this figure.
The numerical error introduced by our method increases weakly for larger
datasets $\sim N^{0.15}$ and roughly linearly with $J$.

@@

We continue by measuring the real-world computational efficiency of this
method as a function of $N$ and $J$.
This experiment was run on a 2013 MacBook Pro with a dual 2.6 GHz Intel Core
i5 processor, but find similar performance on other platforms.
The \celerite\ implementation used for this test is the Python interface with
a C++ back end, but there is little overhead introduced by Python so we find
similar performance using implementations in C++ and Julia
\citep{Bezanzon:2012}.
\Figure{benchmark} shows the computational cost of evaluating
\eq{gp-likelihood} using \celerite\ as a function of $N$ and $J$.
The empirical scaling as a function of data size matches the theoretical
scaling of $\mathcal{O}(N)$ for all values of $N$ and the scaling with the
number of terms is the theoretical $\mathcal{O}(J^2)$ for $J \gtrsim 10$, with
some overhead for smaller models.
For comparison, the computational cost of the same computation using the Intel
Math Kernel Library, a highly optimized general dense linear algebra package,
is shown as a dashed line.
\celerite\ is faster than the reference implementation for all but the
smallest datasets with many terms.

@@

\response{
To show the application of \celerite, we use it to perform posterior
inference in two examples with realistic simulated data and three examples of
real-world research problems.
These examples have been chosen to justify the application of \celerite\ for a
range of use cases, but they are by no means exhaustive.
These examples are framed in the time domain with a clear bias in favor of
large homogeneous photometric surveys, but \celerite\ can also be used for
other one-dimensional problems like, for example, spectroscopy, where
wavelength~--~instead of time~--~is the independent coordinate \citep[see][for
a potential application]{Czekala:2017}.

@@

We then sample the joint posterior probability (\eqalt{joint-post}) using
\emcee\ \citep{Goodman:2010, Foreman-Mackey:2013}.
We initialize \exampleinwalkers~walkers in a Gaussian ball with width
$10^{-4}$ in each parameter around the MAP parameter vector, run
\exampleinburn~steps of burn-in, and \exampleinsteps~steps of MCMC.
To assess convergence, we estimate the mean integrated autocorrelation time of
the chain across parameters \citep{Sokal:1989, Goodman:2010} and find that the
chain results in \exampleineff~effective samples.

Each sample in the chain corresponds to a model PSD and we compare this
posterior constraint on the PSD to the true spectral density in the
right-hand panel of \Figure{simulated-correct}.
In this figure, the true PSD is plotted as a dashed black line and the
numerical estimate of the posterior inference is shown as a blue contour
indicating 68\% of the MCMC samples.
It is clear from this figure that, as expected, the inference correctly
reproduces the true PSD.
}

@@

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=0.8\textwidth]{figures/rotation/rotation.pdf}
\caption{Inferred constraints on a quasiperiodic GP model using the covariance
    function in \eq{rot-kernel} and two quarters of \kepler\ data.
(top left) The \kepler\ data (black points) and the maximum likelihood model
    prediction (blue curve) for a 60~day subset of the data used.
    The solid blue line shows the predictive mean and the blue contours show
    the predictive standard deviation.
(bottom left) The residuals between the mean predictive model and the data
    shown in the top left figure.
(right) The posterior constraint on the rotation period of KIC~1430163 using
    the dataset and model from \Figure{rotation}.
    The period is the parameter $P_\mathrm{rot}$ in \eq{rot-kernel} and this
    figure shows the posterior distribution marginalized over all other
    nuisance parameters in \eq{rot-kernel}.  This is consistent with the
    published rotation period made using the full
    \kepler\ baseline shown as a vertical gray line \citep{Mathur:2014}.
    \figurelabel{rotation}}
\end{center}
\end{figure}

@@

\response{
For this demonstration, we randomly select a month-long segment of PDC data
\citep{Stumpe:2012, Smith:2012} \kepler\ data.
Unlike the previous examples, the posterior distribution is sharply
multimodal and na\"ively maximizing the posterior using \project{L-BFGS-B} is
not practical.
Instead, we start by estimating the initial values for $\nu_\mathrm{max}$ and
$\Delta\nu$ using only the month-long subset of data and following the
standard procedure described by \citet{Huber:2009}.
We then run set of \project{L-BFGS-B} optimizations with values of
$\nu_\mathrm{max}$ selected in logarithmic grid centered on our initial
estimate of $\nu_\mathrm{max}$ and initial values of $\Delta\nu$ computed
using the empirical relationship between these two quantities
\citep{Stello:2009}.
This initialization procedure is sufficient for this example, but the general
application of this method will require a more sophisticated prescription.

@@

\begin{figure}[!htbp]
\begin{center}
\includegraphics[width=0.8\textwidth]{figures/astero/astero.pdf}
\caption{
\response{
(top) The \kepler\ data (black points) and the maximum likelihood model
    prediction (blue curve) for a 10~day subset of the month-long dataset that
    was used for the fit.
    The solid blue line shows the predictive mean and the blue contours show
    the predictive standard deviation.
(bottom) The residuals between the mean predictive model and the data
    shown in the top figure.
}
    \figurelabel{astero}}
\end{center}
\end{figure}

>>>>

The growing field of large-scale time domain astronomy requires methods for
probabilistic data analysis that are computationally tractable, even with
large datasets.
Gaussian Processes are a popular class of models used for this purpose but,
since the computational cost scales as the cube of the number of data points,
their application has been limited to relatively small datasets.
In this paper, we present a novel method for Gaussian Process modeling in
one-dimension where the computational requirements scale linearly with the
size of the dataset.
We demonstrate the method by applying it to simulated and real astronomical
time series datasets.
These demonstrations are examples of probabilistic inference of stellar
rotation periods, asteroseismic oscillation spectra, and transiting planet
parameters.
The method exploits structure in the problem when the covariance function is
expressed as a mixture of complex exponentials, without requiring evenly
spaced observations or uniform noise.
This form of covariance arises naturally when the process is a mixture of
stochastically-driven damped harmonic oscillators~--~providing a physical
motivation for and interpretation of this choice~--~but we also demonstrate
that it is effective in other cases.
We present a mathematical description of the method and compare it to existing
scalable Gaussian Process methods.
The method is flexible, fast, and most importantly, interpretable, with a wide
range of potential applications within astronomical data analysis and beyond.
We provide well-tested and documented open-source implementations of this
method in \project{C++}, \project{Python}, and \project{Julia}.

@@

We now turn to a discussion of \celeriteterm\ as a model of astrophysical
variability.
A common concern in the context of GP modeling in astrophysics is the lack of
physical motivation for the choice of kernel functions.
Kernels are often chosen simply because they are popular, with little
consideration of the impact of this decision.
In this section, we discuss an exact physical interpretation of the
\celeriteterm\ kernel that is applicable to many astrophysical systems, but
especially the time-variability of stars.

@@

For these definitions, the kernel is
\begin{equation}\eqlabel{sho-kernel}
k_\mathrm{SHO}(\tau;\,S_0,\,Q,\,\omega_0) =
S_0\,\omega_0\,Q\,e^{-\frac{\omega_0\,\tau}{2Q}}\,
\begin{cases}
    \cosh{(\eta\,\omega_0\,\tau)} +
        \frac{1}{2\,\eta\,Q}\,\sinh{(\eta\,\omega_0\,\tau)}, & 0 < Q < 1/2\\
    2\,(1+\omega_0\,\tau), & Q = 1/2\\
    \cos{(\eta\,\omega_0\,\tau)} +
        \frac{1}{2\,\eta\,Q} \sin{(\eta\,\omega_0\,\tau)},& 1/2 < Q\\
\end{cases}
\end{equation}
where $\eta = \vert 1-(4\,Q^2)^{-1}\vert^{1/2}$.
Because of the damping, the characteristic oscillation frequency in this
model, $d_j$, for any finite quality factor $Q > 1/2$, is not equal to the
frequency of the undamped oscillator, $\omega_0$.

@@

{\item For $Q = 1/\sqrt{2}$, \eq{sho-psd} simplifies to
\begin{eqnarray}\eqlabel{granulation-psd}
S(\omega) = \sqrt{\frac{2}{\pi}}\,\frac{S_0}{(\omega/\omega_0)^4+1} \quad.
\end{eqnarray}
This functional form is commonly used to model for the background granulation
noise in asteoreseismic and helioseismic analyses \citep{Harvey:1985, Michel:2009,
Kallinger:2014}.
The kernel for this value of $Q$ is
\begin{equation}
k(\tau) = S_0\,\omega_0\,e^{-\frac{1}{\sqrt{2}}\,\omega_0\,\tau}\,
    \cos{\left(\frac{\omega_0\,\tau}{\sqrt{2}}-\frac{\pi}{4}\right)} \quad.
\end{equation}}

{\item Substituting $Q = 1/2$, \eq{sho-psd} becomes
\begin{eqnarray}
S(\omega) =
    \sqrt{\frac{2}{\pi}}\,\frac{S_0}{\left[(\omega/\omega_0)^2+1\right]^2}
\end{eqnarray}
with the corresponding covariance function (using
\eqalt{celerite-kernel} and \eqalt{sho-real})
\begin{eqnarray}\eqlabel{approx-matern}
k(\tau) &=& \lim_{f \to 0}\,
    \frac{1}{2}\,S_0\,\omega_0\,
    \left[\left(1+1/f\right)\,e^{-\omega_0\,(1-f)\,\tau} +
          \left(1-1/f\right)\,e^{-\omega_0\,(1+f)\,\tau}
    \right] \\
&=& S_0\,\omega_0\,e^{-\omega_0\,\tau}\,[1+\omega_0\,\tau]
\end{eqnarray}
or, equivalently (using \eqalt{celerite-kernel} and \eqalt{sho-complex})
\begin{eqnarray}\eqlabel{approx-matern2}
k(\tau) &=& \lim_{f \to 0}\,
    S_0\,\omega_0\,e^{-\omega_0\,\tau}\,
    \left[\cos(f\,\tau) + \frac{\omega_0}{f}\,\sin(f\,\tau)\right] \\
&=& S_0\,\omega_0\,e^{-\omega_0\,\tau}\,[1+\omega_0\,\tau] \quad.
\end{eqnarray}
This covariance function is also known as the Mat\'ern-3/2 function
\citep{Rasmussen:2006}.
This suggests that the Mat\'ern-3/2 covariance can be approximated using
the \celeriteterm\ framework with a small value of $f$ in \eq{approx-matern2} but we
caution that this might lead to numerical issues for the solver.
}

@@

The relationship between the \celeriteterm\ model and semiseparable linear
algebra was first recognized by Ambikasaran \citep{Ambikasaran:2015} building
on earlier work by Rybicki \& Press \citep{Rybicki:1995}.
Ambikasaran derived a scalable direct solver for any general semiseparable
matrix and applied this method to \celeriteterm\ models with $b_j = 0$ and
$d_j = 0$.
During the preparation of this paper, we generalized this earlier method to
include non-zero $b_j$ and $d_j$, but it turns out that we can derive a higher
performance algorithm by restricting our method to positive definite matrices.
In \app{psd}, we discuss methods for ensuring that a \celeriteterm\ model is
positive.

@@

\subsection{Cholesky factorization of positive definite semiseparable matrices}

For our purposes, we want to compute the log-determinant of $K$ and solve
linear systems of the form $K\,\bvec{z}=\bvec{y}$ for $\bvec{z}$, where $K$ is
a rank-$R$ semiseparable positive definite matrix.
These can both be computed using the Cholesky factorization of $K$:
\begin{eqnarray}
K = L\,D\,L^\T
\end{eqnarray}
where $L$ is a lower triangular matrix with unit diagonal and $D$ is a
diagonal matrix.
We use the ansatz that $L$ has the form
\begin{eqnarray}
    L = I + \mathrm{tril} (U\,W^T)
\end{eqnarray}
where $I$ is the identity, $U$ is the matrix from above, and $W$ is an unknown
$N \times R$ matrix.
Combining this with \eq{semi-sep}, we find the equation
\begin{eqnarray}\eqlabel{cholesky}
A + \mathrm{tril}(U\,V^\T) + \mathrm{triu}(V\,U^\T)
&=& \left[I + \mathrm{tril} (U\,W^T)\right]\,D\,
    {\left[I + \mathrm{tril} (U\,W^T)\right]}^\T \\
&=& D + \mathrm{tril}(U\,W^\T)\,D
    + D\,\mathrm{triu}(W\,U^\T) \nonumber\\
&& +\, \mathrm{tril}(U\,W^\T)\,D\,\mathrm{triu}(W\,U^\T) \nonumber
\end{eqnarray}
that we can solve for $W$.
Setting each element on the left side of \eq{cholesky} equal to the
corresponding element on the right side, we derive the following recursion
relations
\begin{eqnarray}
S_{n,j,k} &=& S_{n-1,j,k} + D_{n-1,n-1}\,W_{n-1,j}\,W_{n-1,k} \nonumber\\
D_{n,n} &=& A_{n,n} -
    \sum_{j=1}^{R}\sum_{k=1}^{R} U_{n,j}\,S_{n,j,k}\,U_{n,k}
    \nonumber\\
W_{n,j} &=& \frac{1}{D_{n,n}}\left[ V_{n,j} -
    \sum_{k=1}^{R}U_{n,k}\,S_{n,j,k} \right]
    \eqlabel{algorithm}
\end{eqnarray}
where $S_n$ is a symmetric $R \times R$ matrix and every element
$S_{1,j,k}=0$.
The computational cost of one iteration of the update in \eq{algorithm} is
$\mathcal{O}(R^2)$ and this must be run $N$ times~--~once for each row~--~so
the total cost of this factorization scales as $\mathcal{O}(N\,R^2)$.

After computing the Cholesky factorization of this matrix $K$, we can also
apply its inverse and compute its log-determinant in $\mathcal{O}(N\,R)$ and
$\mathcal{O}(N)$ operations respectively.
The log-determinant of $K$ is given by
\begin{eqnarray}
    \ln \det K = \sum_{n=1}^N \ln D_{n,n}
\end{eqnarray}
where, since $K$ is positive definite, $D_{n,n}>0$ for all $n$.

@@

The algorithm derived here is a general method for factorizing, solving, and
computing the determinant of positive definite rank-$R$ semiseparable
matrices.
As we will show below, this method can be used to compute
\eq{gp-likelihood} for \celeriteterm\ models.
In practice, when $K$ is positive definite and this method is applicable, we
find that this method is about a factor of 20 faster than an optimized
implementation of the earlier, more general method from
\citet{Ambikasaran:2015}.

@@

First, we empirically confirm that the method solves systems of the form
$K\,\bvec{z} = \bvec{y}$ to high numerical accuracy.
The left panel of \Figure{error} shows the L-infinity norm for the residual
$\bvec{z} - K^{-1}\,\bvec{y}$ for a range of dataset sizes $N$ and numbers of
terms $J$.
For each pair of $N$ and $J$ we repeated the experiment 10 times with
different randomly generated datasets and \celeriteterm\ parameters, and
averaged across experiments to remove some sampling noise from this figure.
The numerical error introduced by our method increases weakly for larger
datasets $\sim N^{0.15}$ and roughly linearly with $J$.

@@

We continue by measuring the real-world computational efficiency of this
method as a function of $N$ and $J$.
This experiment was run on a 2013 MacBook Pro with a dual 2.6 GHz Intel Core
i5 processor, but we find similar performance on other platforms.
The \celerite\ implementation used for this test is the \project{Python}
interface with a \project{C++} back end, but there is little overhead
introduced by \project{Python} so we find similar performance using
implementations in \project{C++} and \project{Julia} \citep{Bezanzon:2012}.
\Figure{benchmark} shows the computational cost of evaluating
\eq{gp-likelihood} using \celerite\ as a function of $N$ and $J$.
The empirical scaling as a function of data size matches the theoretical
scaling of $\mathcal{O}(N)$ for all values of $N$ and the scaling with the
number of terms is the theoretical $\mathcal{O}(J^2)$ for $J \gtrsim 10$, with
some overhead for smaller models.
For comparison, the computational cost of the same computation using the Intel
Math Kernel Library, a highly optimized general dense linear algebra package,
is shown as a dashed line.
\celerite\ is faster than the reference implementation for all but the
smallest datasets with many terms.

@@

\response{
To demonstrate the application of \celerite, we use it to perform posterior
inference in two examples with realistic simulated data and three examples of
real-world research problems.
These examples have been chosen to justify the use of \celerite\ for a range
of research problems, but they are by no means exhaustive.
These examples are framed in the time domain with a clear bias in favor of
large homogeneous photometric surveys, but \celerite\ can also be used for
other one-dimensional problems like, for example, spectroscopy, where
wavelength~--~instead of time~--~is the independent coordinate \citep[see][for
a potential application]{Czekala:2017}.

@@

We then sample the joint posterior probability (\eqalt{joint-post}) using
\emcee\ \citep{Goodman:2010, Foreman-Mackey:2013}.
We initialize \exampleinwalkers~walkers by sampling from three-dimensional
isotropic Gaussian centered the MAP parameter vector and with a standard
deviation of $10^{-4}$ in each dimension.
We then run \exampleinburn~steps of burn-in and \exampleinsteps~steps of MCMC.
To assess convergence, we estimate the mean integrated autocorrelation time of
the chain across parameters \citep{Sokal:1989, Goodman:2010} and find that the
chain results in \exampleineff~effective samples.

Each sample in the chain corresponds to a model PSD and we compare this
posterior constraint on the PSD to the true spectral density in the
right-hand panel of \Figure{simulated-correct}.
In this figure, the true PSD is plotted as a dashed black line and the
numerical estimate of the posterior constraint on the PSD is shown as a blue
contour indicating 68\% of the MCMC samples.
It is clear from this figure that, as expected, the inference correctly
reproduces the true PSD.
}

@@

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=0.9\textwidth]{figures/rotation/rotation.pdf}
\caption{Inferred constraints on a quasiperiodic GP model using the covariance
    function in \eq{rot-kernel} and two quarters of \kepler\ data.
(top left) The \kepler\ data (black points) and the maximum likelihood model
    prediction (blue curve) for a 60~day subset of the data used.
    The solid blue line shows the predictive mean and the blue contours show
    the predictive standard deviation.
(bottom left) The residuals between the mean predictive model and the data
    shown in the top left figure.
(right) The posterior constraint on the rotation period of KIC~1430163 using
    the dataset and model from \Figure{rotation}.
    The period is the parameter $P_\mathrm{rot}$ in \eq{rot-kernel} and this
    figure shows the posterior distribution marginalized over all other
    nuisance parameters in \eq{rot-kernel}.  This is consistent with the
    published rotation period made using the full
    \kepler\ baseline shown as a vertical gray line \citep{Mathur:2014}.
    \figurelabel{rotation}}
\end{center}
\end{figure}

@@

\response{For this demonstration, we randomly select a month-long segment of
PDC \kepler\ data \citep{Stumpe:2012, Smith:2012}.
Unlike the previous examples, the posterior distribution is sharply
multimodal and na\"ively maximizing the posterior using \project{L-BFGS-B} is
not practical.
Instead, we start by estimating the initial values for $\nu_\mathrm{max}$ and
$\Delta\nu$ using only the month-long subset of data and following the
standard procedure described by \citet{Huber:2009}.
We then run set of \project{L-BFGS-B} optimizations with values of
$\nu_\mathrm{max}$ selected in logarithmic grid centered on our initial
estimate of $\nu_\mathrm{max}$ and initial values of $\Delta\nu$ computed
using the empirical relationship between these two quantities
\citep{Stello:2009}.
This initialization procedure is sufficient for this example, but the general
application of this method will require a more sophisticated prescription.

@@

\begin{figure}[!htbp]
\begin{center}
\includegraphics[width=0.5\textwidth]{figures/astero/astero.pdf}
\caption{
\response{
(top) The \kepler\ data (black points) and the maximum likelihood model
    prediction (blue curve) for a 10~day subset of the month-long dataset that
    was used for the fit.
    The solid blue line shows the predictive mean and the blue contours show
    the predictive standard deviation.
(bottom) The residuals between the mean predictive model and the data
    shown in the top figure.
}
    \figurelabel{astero}}
\end{center}
\end{figure}
